Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-07-31 anzjs2018/sections/02_gtfs.tex
make a bunch of changes

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who use it in Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS-realtime},
which is only available in a subset of the providers due to the requirement of 
on board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we describe how we constructed a transit road network consisting of intersections
and the roads between them.
Generalising to physical roads removes route dependencies,
which increases the quantity of data available for each road
and allows each route to be represented as a sequence of road segments,
which have individual travel time distributions (Section~\ref{sec:kf}).

The transit network was constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road,
by using bus stops as nodes in the network
and the connecting roads as edges
(Figure~\ref{fig:network_creation}).
In this way, routes that service the same subsequence of stops
all contribute to congestion information for the connecting roads.
Although there are several drawbacks to this method,
such as road segment overlaps where routes merge between stops,
this is minimal and our approach is sufficient for the current work.
Future work will investigate methods of detecting where routes merge
(physical intersections) and using these as nodes in the network.

>>>>

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who use it in Google Maps Transit Directions,
it is used by over 1100~transit providers around the world\footnote{%
source: \url{http://transitfeeds.com} as of August 2019},
including here in Auckland, New Zealand.
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item and \emph{shapes}, the sequence of points defining a vehicle's path along a route.
\end{itemize}
The second component is \emph{GTFS-realtime},
which is only available in a subset of the providers due to the requirement of 
on board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable, at least directly.
In order to compute travel times efficiently
and make use of all available data pooled across multiple routes,
we constructed a \emph{road network},
allowing us to model each physical road as a unique entity.
In this way, every bus travelling along a road contributes to its state,
and can be used to predice travel times for all upcoming buses 
travelling along the road,
irrespective of the route they are servicing.

The transit network was constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road,
by using bus stops as nodes in the network
and the connecting roads as edges
(Figure~\ref{fig:network_creation}).
In this way, routes that service the same subsequence of stops
all contribute to congestion information for the connecting roads.
Although there are several drawbacks to this method,
such as road segment overlaps where routes merge between stops,
this is minimal and our approach is sufficient for the current work.

### 2019-07-31 anzjs2018/sections/03_models.tex
make a bunch of changes

Real-time tracking applications often use
a Bayesian filtering approach,
or recursive Bayesian estimation,
in which the data is processed sequentially instead of all at once,
reducing computational demands
\citep{Anderson_2012}.
These models assume an underlying Markov process with state $\mathcal{X}_k$
and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ describes the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector
$\bX_k = \left[x_k\ \dot x_k\ \bz\right]^\top$.
% \begin{equation*}
% \label{eq:vehicle_state}
% \bX_k = 
% \begin{bmatrix}
%     x_k \\ \dot x_k \\ \bz
% \end{bmatrix}.
% \end{equation*}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$
% \begin{equation}
% \label{eq:vehicle_obs}
% \bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
% \end{equation}

We use a particle filter to estimate $\bX_k$,
due to its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation*}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation*}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation*}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation*}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation*}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
\end{equation*}

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

>>>>

Real-time tracking applications often use
a Bayesian filtering approach,
or recursive Bayesian estimation,
in which the data is processed sequentially instead of all at once,
reducing computational demands
\citep{Anderson_2012}.
These models assume an underlying Markov process with state $\boldsymbol{\mathcal{X}}_k$,
an $n\times1$ column vector,
and system noise $\boldsymbol{\omega}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{Q})$ 
of which observations $\boldsymbol{\mathcal{Y}}_k$,
an $m\times1$ column vector, are made
with error $\boldsymbol{\nu}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\boldsymbol{\mathcal{X}}_k &= f(\boldsymbol{\mathcal{X}}_{k-1}, \boldsymbol{\omega}_k) \\
\boldsymbol{\mathcal{Y}}_k &= h(\boldsymbol{\mathcal{X}}_k) + \boldsymbol{\nu}_k
\end{split}
\end{equation}
The transition function $f:\mathbb{R}^n\mapsto\mathbb{R}^n$ 
describes the relationship between consecutive states,
while the measurement function $h:\mathbb{R}^n\mapsto\mathbb{R}^m$ is a deterministic function
mapping the underlying state to the observable state.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
% The travel times along road segments, $\bz = (z_1,\ldots,z_L)^\top$, in seconds,
% are estimated sequentially as the vehicle traverses the route,
% and are not included explicitely in the state.
These are combined into the state vector
$\bX_k = \left[x_k\ \dot x_k\right]^\top$,
of which observations $\bY_k$ are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.

We use a particle filter to estimate $\bX_k$,
due to its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015},
as well as providing an easy way to estimate travel times
$\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
as the vehicle traverses the route.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in Figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is that the likelihood function is intuitively based
on the physical distance between the vehicle observation and the state's
estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}),
rather than the reverse of first estimating the ``observed'' distance travelled.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$.
The state can then be expressed using the Dirac delta function $\dot\delta_X(x)$ \citep{cn},
so that
\begin{equation*}
p(\bX_{k-1} | \bY_{1:k-1}) \approx 
    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k-1}}(\bX_{k-1}^{(i)})
\end{equation*}
each of which is independently updated or \emph{mutated} using the transition function $f$ (Figure~\ref{fig:pf_state_predict})
\begin{equation*}
p(\bX_k | \bX_{k-1}) \approx 
    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k}}(f(\bX_{k-1}^{(i)}, \psi))
\end{equation*}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation*}
p(\bX_k | \bY_k) \approx  
    \sum_{i=1}^N W_{k} \dot\delta_{\bX_{k}}(\bX_{k}^{(i)})
\end{equation*}

One problem with particle filters is degeneration,
which occurs when the particle sample no longer approximates the target distribution
due to the particles becoming too diverse
and all of the weight falling onto only a small number of them.
This occurs naturally over time, but the rate at which it occurs depends 
on a relationship between how well the transition function predicts future states,
and the variance of those predictions (system noise).
To avoid degeneration,
particle filters use \emph{importance resampling}
using importance weights $\{W_k^{(1)}, \ldots, W_k^{(N)}\}$,
which removes improbable particles and replaces them with more probable ones,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
However, resampling requires sorting $N$ particles,
which is of $\mathcal{O}(N\log N)$ complexity,
so we can reduce computational costs by only resampling when required.
To prevent degeneration of the particle filter state,
the sample is resampled whenever the
effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
falls below a specified threshold $N_{\text{thres}}$.
We used a fixed threshold value of $N_{\text{thres}} = N/10$
for the current work.
\textbf{Check this}.
